---
title: "Reproductible Research, Project 1"
author: "Irving Avila"
date: "03/20/2016"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

##Loading and preprocessing the data


1. Load the data

```{r readData}
activity <- read.csv("activity.csv")
summary(activity)
```


##What is mean total number of steps taken per day?

Missing values in the dataset are ignored.

1. Total number of steps taken per day

```{r steps}
#Calculate the total number of steps per day.
spd <- tapply(activity$steps, activity$date, sum, na.rm = TRUE)
```

2. Histogram of the total number of steps taken each day

```{r histogram, echo = FALSE}
barplot(spd, las = 2, space = 0, cex.axis = .8, cex.names = 0.5, xlab = "Date", ylab = "Steps")
```

3. Mean and median of the total number of steps taken per day

```{r meanmedian}
mean <- mean(spd)
median <- median(spd)
```
* Mean: `r mean`
* Median: `r median`

##What is the average daily activity pattern?

1. Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged accros all days (y-axis)

```{r plot}
spd1 <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
```

```{r plot1, echo = FALSE}
plot(names(spd1), spd1, type = "l", xlab = "Interval", ylab = "Average steps over all days")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r maxsteps}
max <- names(which.max(spd1))
max
```
The maximum value is in the interval `r max`. 


##Imputing missing values

There are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations of the data.

1. Total number of missing values in the dataset

```{r missing}
missing <- sum(is.na(activity$steps))
```
The number of missing values is `r missing`.

2. The missing values can be filled using the mean/median of the day on which the missing value appears

```{r replace}
#Check to see if there are missing values
for (i in 1:length(activity$steps)) {
    #if there is one, it is replaced with the mean of the day divided by the total of intervals.
    if (is.na(activity$steps[i])) {
        activity$steps[i] <-  spd[activity$date[i]] / length(activity$steps[activity$date == activity$date[i]])
    }
}
```

3. New dataset, histogram, mean and median

```{r newset}
spd2 <- tapply(activity$steps, activity$date, sum)
```

```{r hist2, echo = FALSE}
barplot(spd2, las = 2, space = 0, cex.axis = .8, cex.names = 0.5, xlab = "Date", ylab = "Steps (with NAs replaced)")
options(scipen=999)
```

```{r meanmedian2}
mean <- mean(spd2)
median <- median(spd2)
```

* Mean with NAs replaced: `r mean`
* Median with NAs replaced: `r median`

The mean and median remain the same, because NA values were replaced with the mean.

4. Histogram of the total number of steps taken each day.

##Are there differences in activity patterns between weekdays and weekends?

1. Creating a  new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r week}
#Change the locale to english so that the weekdays function works (my system language isn't english)
Sys.setlocale("LC_ALL","English")
#Change dates from factor to Date
activity$date <- as.Date(activity$date)
#Add a factor variable for weekday and weekend
activity$weekday <- as.factor(rep("weekday",length(activity$date)))
levels(activity$weekday) = c("weekday","weekend")
activity$weekday[weekdays(activity$date) == "Saturday" | weekdays(activity$date) == "Sunday"] <- "weekend"

```

2. Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r plot2}
#Calculate the mean for every interval
aggdata <- with(activity, aggregate(steps, by = list(interval, weekday), FUN = mean))
par(mfrow = c(2,1))
#Weekday plot
plot(unique(aggdata$Group.1), aggdata$x[aggdata$Group.2 == "weekday"], xlab = "Interval", ylab = "Average steps", main = "Weekdays", type = "l")
#Weekend plot
plot(x = unique(aggdata$Group.1), y = aggdata$x[aggdata$Group.2 == "weekend"], xlab = "Interval", ylab = "Average steps", main = "Weekends", type = "l")
```
